---
output: github_document
---


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(tidyverse)
library(knitr)
library(dataDocumentation)
library(knitr)
```

# dataDocumentation


The goal of the **dataDocumentation** package is to make documenting your data easier in the Between the Fjords lab.
The goal is to make openly available for the members in the BTF lab and also beyond.
For this, we use reproducible and transparent workflows and properly document each step and object.

Each dataset should be properly documented with a **data dictionary**.
A data dictionary is a document that describes each variable of a dataset.
This package helps you to create a data dictionary for all your datasets.

Another aspect of our workflow is to make different datasets talk to each other.
For this, we need a pre-defined **naming convention**.
Therefore, this package contains functions to create the **meta data** for the Funder and ThreeD project.
The meta data contains a complete file with the experimental strucutre including site, block, plot, treatment names.


## Installation

You can install `dataDocumentation` from  [GitHub](https://github.com/) with:

``` {r install, eval=FALSE, echo=TRUE}
# if needed install the remotes package
install.packages("remotes")

# then install the dataDocumentation package
remotes::install_github("audhalbritter/dataDocumentation")

# and load it
library(dataDocumentation)

```

## Data dictionaries

### description table

The description table contains the information about each variable in your dataset.
You can create this table using the `make_description_table()` function.
By default, this function creates a **data_dic** folder and writes the description table as a CSV file:

``` {r description-table, eval=TRUE, echo=TRUE}

data(biomass)
description <- make_description_table(data = biomass,
                                      table_ID = "biomass",
                                      path = NULL)
description |> 
  knitr::kable()

```

To write the description table to a CSV file (default path is "data_dic"):

``` {r description-table-write, eval=FALSE, echo=TRUE}

# This creates data_dic/description_table.csv
make_description_table(data = biomass,
                       table_ID = "biomass")

# Or specify a custom directory
make_description_table(data = biomass,
                       table_ID = "biomass",
                       path = "my_custom_folder")

```

The function automatically extracts all variable names from your data and creates a template with the following columns:

**TableID** is a unique identifier for each dataset.
This is useful when you have the same variable name in different datasets with different meanings.
For example, if you use "value" for both biomass and decomposition datasets, you can distinguish them using TableID.

**Variable name** is the name of the variable in the dataset.

**Description** is the description of the variable.
It should be concise but detailed enough to understand the variable.
For example: 

- Year of sampling
- Unique plot ID is a combination of site, block and treatment
- Removed functional group, including forbs, bryophytes, graminoids

**Units** defines the unit of a variable.
For example g, cm^3^

**How measured** describes how the variable was obtained.
For example measured, recorded or defined.

If you have already created description tables for other datasets, you can provide a **previous_description_table** to pre-fill variables that are already documented:

``` {r description-table-previous, eval=TRUE, echo=TRUE}

data(previous_description_table)
previous_description_table |> 
  knitr::kable()

description <- make_description_table(data = biomass,
                                      table_ID = "biomass",
                                      previous_description_table = previous_description_table,
                                      path = NULL)
description |> 
  knitr::kable()

```

This way, common variables like `year` and `siteID` are automatically filled in, and you only need to add descriptions for the new variables.


### AI-powered description generation (optional)

**NEW!** You can now use AI to automatically generate descriptions for your variables using Google's Gemini API.

First, install the `gemini.R` package and set up your API key:

``` {r ai-setup, eval=FALSE, echo=TRUE}

# Install gemini.R package
install.packages("gemini.R")

# Get a free API key at: https://aistudio.google.com/apikey
# Then set it as an environment variable
Sys.setenv(GEMINI_API_KEY = "your-api-key-here")

```

Now you can use `populate_description_table()` to automatically fill in descriptions:

``` {r ai-populate, eval=FALSE, echo=TRUE}

# Create empty description table
description <- make_description_table(data = biomass,
                                      table_ID = "biomass",
                                      path = NULL)

# Automatically populate with AI
description_filled <- populate_description_table(
  data = biomass,
  description_table = description
)

# Review and edit the results
description_filled |> 
  knitr::kable()

```

You can provide additional context to help the AI generate better descriptions:

``` {r ai-context, eval=FALSE, echo=TRUE}

description_filled <- populate_description_table(
  data = biomass,
  description_table = description,
  dataset_context = "Biomass data from functional group removal experiment in alpine sites in western Norway"
)

```

The AI will analyze your data and generate:

- **Description**: A concise one-sentence description of each variable
- **Units**: The unit of measurement (if applicable)
- **How measured**: Whether the variable was defined, recorded, or measured

**Important notes:**

- The function only fills in variables that have NA values, so you can run it multiple times
- Always review and edit the AI-generated descriptions before using them
- The function uses the first 100 rows of your data for context


### make data dictionary

The next step is to make the data dictionary.
For this you can use the `make_data_dic()` function.


``` {r data-dic, eval=TRUE, echo=TRUE}

# load dataset and description table
data(description_table)
data(biomass)
data_dic <- make_data_dictionary(data = biomass,
                                 description_table = description_table,
                                 table_ID = "biomass",
                                 keep_table_ID = FALSE)

```

This function gets the **range**, and **class** for each variable in your dataset and combines this information with the **description table**.
The output looks like this:



``` {r data-dic-output, eval=TRUE, echo=FALSE}

data_dic |> 
  kable()

```



## Metadata

The dataDocumentation also provides functions to create metadata for FUNDER, Durin and ThreeD.
The functions are called `create_funder_meta_data()`, `create_durin_meta_data()` and `create_threed_meta_data()`.
These functions creates a tibble with the complete meta data including siteID, blockID, treatment and plotID for FUNDER.
For Durin there it can create 3 different meta data tables (see below).
Origin and destination siteID, blockID and plotID, turfID, warming, grazing and Nlevel for ThreeD.

Just run the code like this:

``` {r meta-data, eval=TRUE, echo=TRUE}

meta_data <- create_funder_meta_data()

meta_data

```

If you want to save the metadata as a csv file, set `csv_output = TRUE` and you can choose a file name `filename` (not mendatory).
The data will be stored in the project  directory.
The file name always contains the project name (FUNDEr, Durin or ThreeD).

``` {r meta-data-output, eval=FALSE, echo=TRUE}

meta_data <- create_funder_meta_data(csv_output = TRUE, filename = "biomass")

```

### Durin meta data

For the Durin project, the meta data can be made for the 4Corners, DroughtNet or Nutrient study.
The `create_durin_meta_data()` function has the argument `study`, which has to be defined. *4Corners*: study on the 4 main Durin sites; *DroughtNet*: droughtNet experiment at Lygra and Tjotta; *Nutrient*: nutrient experiment at Lygra


### funcabization

At some sites, FunCaB and FUNDER datasets have different blockID's.
This is problematic, when merging datasets from the different project.
`funcabization()` is a function to solve this problem.
This function can convert FUNDER terminology to FunCaB terminology and vice versa.
Note that data on OSF should be on FunCaB nomencaltur.

`funcabization()` has two arguments.
The first one is `dat` and takes the data for input.
The second argument is `covnert_to` and here you define if you want to convert a dataset to *FunCaB* or *Funder* nomenclature.

``` {r funcabization, eval=FALSE, echo=TRUE}

data(funder)
funcabization(dat = funder, convert_to = "FunCaB")

data(funcab)
funcabization(dat = funcab, convert_to = "Funder")

```
